---
title: "Possible Bug in base::sort()"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, comment="#>")
```

This covers two related cases of incorrect output in `base::sort()` that occur when using partial sorting on vectors with `NA`s. It seems very surprising that there would still be bugs in a prominent base function, so it may be that my understanding is off. However, although `sort()` is an central function, the combination of arguments where the issue occurs will be less frequently used and this could explain why it has not been picked up.

## Background on `sort()`

The optional `partial` argument of `sort()` gives indices of the results which are to be placed in the correct position. Elements at other indices are not guaranteed to be in the correct position. The option to partially sort is included for speed when the position of only a small number of elements is needed (for example, finding the second largest element). From `?sort`: _"for each of the result values in a specified position, any values smaller than that one are guaranteed to have a smaller index in the sorted array and any values which are greater are guaranteed to have a bigger index in the sorted array."_ 

My understanding of partial sorting (although this is not explicit in the help) is that `sort(x, partial = n)[n]` should give the same result as `sort(x)[n]`.

By default, `NA` values in the input vector are removed, but the `na.last` argument gives an option to keep them, putting them last if `na.last=TRUE`, or first if `na.last=FALSE`.

The combination of these two arguments appears to give two kinds of unexpected outputs. The first gives  an error where there should be a valid output, while the second gives output that is incorrect. Both issues have the same cause - any `NA`s are removed before sorting and then attached to the front or end of the result. This means that the partial indices used in the sort are not the same as the indices of the result after `NA`s have been put back. 

## Issue 1: Index out of Bounds Error

The first issue occurs with `na.last = TRUE` when an index in the `partial` argument is greater than the number of non-NA values in the input.

```{r issue_1, error = TRUE}
sort(c(1:3, NA), partial = 4, na.last = TRUE)
sort(c(1, NA, NA, NA), partial = 2, na.last = TRUE)
```

This at least fails safely, but it is not correct to say the index is out of bounds as can be seen from the corresponding non-partial sorts which do have values (albeit `NA`) in the indices corresponding to the previous `partial` arguments.

```{r issue_1b, error = TRUE}
sort(c(1:3, NA), na.last = TRUE)
sort(c(1, NA, NA, NA), na.last = TRUE)
```

The cause is that the partial sort is done on the vector with `NA`s removed.

## Issue 2: Incorrect Output

The second, more serious issue, occurs when `na.last = FALSE`. It does not occur with all inputs which is one reason why it is easy to miss.

```{r issue_2}
x <- c(7, 2, 4, 5, 3, 6, NA)
# Full sort (correct)
sort(x, na.last = FALSE)
# Partial sort
sort(x, partial = 3, na.last = FALSE)
```

The non-partial sort is correct and has a different value in index 3 than the partial sort, where it is incorrect. 

The `NA` is being removed before sorting so that sorting is only guaranteed to be correct for an incorrect index (the `4` in this case which is the third position after the `NA` is removed).

```{r issue_2b}
sort(x, partial = 3)
```

Note that the symmetric mistakes of these two issues with `na.last` and ordering reversed do not occur because the `decreasing=TRUE` argument cannot be used with `partial`.
